---
title: "Analysis of AMP-AD with dream"
subtitle: 'Data from [Wang, et al.](https://www.nature.com/articles/sdata2018185)'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
documentclass: article
output: 
  html_document:
  toc: true
  smart: false
---


```{r initialize, cache=FALSE, echo=FALSE, message=FALSE, results='hide'}
nthreads = 8
# rm(cl)
```

```{r load.always, cache=FALSE, echo=FALSE, message=FALSE, results='hide'}
suppressPackageStartupMessages(library(doParallel))

if( ! exists("cl") ){
	cl <- makeCluster(nthreads)
	registerDoParallel(cl) 
}

suppressPackageStartupMessages(library(synapseClient))

synapseLogin()
```

```{r load.packages, echo=TRUE, message=FALSE, results='hide'}
suppressPackageStartupMessages(library(qvalue))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(variancePartition))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(synapseClient))

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=TRUE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE,
  dev = c("png", "pdf"), 
  fig.width=7, fig.height=7)

options(markdown.HTML.stylesheet = 'css/custom.css')
```

```{r read.geneCounts}
synIds = c('syn16796116', 'syn16796117', 'syn16796121', 'syn16796123')
files = sapply( synIds, function(x) getFileLocation(synGet(x)) )

geneCounts_in = foreach( file = files, .combine=cbind) %do% {
  read.table( file, header=TRUE, row.names=1 )
} 
colnames(geneCounts_in) = gsub("^(\\S+)\\.", "", colnames(geneCounts_in))

# info = fread("/hpc/users/hoffmg01/work/dev_dream/ampad/msbb.good.samples.renamed.tsv")
info = read.table(getFileLocation(synGet('syn16816486')), header=TRUE )
info = data.table(info)

indivs = intersect(info$Lib, colnames(geneCounts_in))
info = info[info$Lib %in% indivs,]
geneCounts = geneCounts_in[,colnames(geneCounts_in) %in% indivs]
```




Meta data is available from https://www.synapse.org/#!Synapse:syn7392158
```{r metadata}
df_rna_covariates = fread(getFileLocation(synGet( 'syn6100548' )))
df_clinical = fread(getFileLocation(synGet( 'syn6101474' )))
df_wes = fread(getFileLocation(synGet( 'syn6101472' )))
sampleInfo = fread(getFileLocation(synGet( 'syn12178047' )))


idx = match(c('BrainID (final)', 'BrainID by annotation'),colnames(sampleInfo))

metadata = merge(info, sampleInfo[Datatype=='RNA-seq',-idx,with=FALSE], by.x="Lib", by.y='Sampleid')

metadata2 = merge(metadata, df_rna_covariates[grep(".bam$", fileName),], by.x="Lib", by.y='sampleIdentifier')

metadata3 = merge(metadata2, df_clinical, by='individualIdentifier')

# when samples were seqenced in multiple batches, 
# take the one with the largest RIN
metadata = metadata3[,.SD[which.max(RIN),],by='Lib']

# keep only OKay
metadata = metadata[(Action=="OKay") & (Region %in% c('BM10', 'BM22', 'BM36', 'BM44')),]

# convert NP1 to disease class
# 1= control
# 2=definite AD
# 3=possible AD
# 4=probable AD
# so control is 1
# and AD is 2-4
Dx = metadata$NP.1 
Dx[Dx==1] = "control"
Dx[Dx==2] = "AD"
Dx[!(Dx %in% c("control", "AD"))] = NA
metadata$DiseaseClass = factor(Dx, c("control", "AD"))

# only individuals with 4 samples, and Braak state and Apo genotype
i1 = table(metadata$individualIdentifier)
idx = metadata[,(individualIdentifier %in% names(i1[i1 ==4])) &  (!is.na(Apo1)) & (!is.na(bbscore)) &(!is.na(DiseaseClass)) &(`Race (final)` %in% c('White')) ]

metadata = metadata[idx,]

# order metadata to batch geneCounts
jointID = intersect(colnames(geneCounts), metadata$Lib)

geneCounts = geneCounts[,colnames(geneCounts) %in% jointID]
metadata = metadata[Lib %in% jointID,]

idx = match(colnames(geneCounts), metadata$Lib)
metadata = metadata[idx,]

identical(colnames(geneCounts), metadata$Lib)

# Age as numeric
metadata$Age = as.numeric(gsub("\\+", "", metadata$AOD))
metadata[,mappingRate := Mapped / TotalReads]
```


```{r RNA.seq_processing}
isexpr = rowSums(cpm(geneCounts)>1) >= 0.1*ncol(geneCounts)
isexpr['ENSG00000104356'] = FALSE # very inconsistent bewteen hg19 and hg38

# Standard usage of limma/voom
genes = DGEList( geneCounts[isexpr,] ) 
genes = calcNormFactors( genes)

design = model.matrix( ~ Region + LIMS , metadata)
vobj_tmp = voom( genes, design, plot=TRUE)

# apply duplicateCorrelation 
dupcor <- duplicateCorrelation(vobj_tmp,design,block=metadata$individualIdentifier)

# run voom considering the duplicateCorrelation results
# in order to compute more accurate precision weights
# Otherwise, use the results from the first voom run
vobj = voom( genes, design, plot=FALSE, block=metadata$individualIdentifier, correlation=dupcor$consensus)
```

```{r define}
synapseLogin()
# load("enrich.RDATA")
load(getFileLocation(synGet( 'syn16816471' )))

# exlucde some gensets: mSigDB C4
geneSetsCombined = geneSetsCombined[grep("^c4", names(geneSetsCombined), invert=TRUE)]
geneSetsCombined = geneSetsCombined[grep("^CMC", names(geneSetsCombined), invert=TRUE)]

# geneInfo = readRDS("geneInfo.RDS")
geneInfo = readRDS(getFileLocation(synGet( 'syn16816472' )))

get_enrich_order = function( res ){
  res$qvalue = qvalue( res$P.Value )$qvalue
  rownames(res) = gsub("\\..*$", "", rownames(res))
  res$gene = geneInfo$geneName[match(rownames(res), geneInfo$Geneid)]
  res$symbol = sub("^(ENSG.*)$", NA, res$gene)

  tstat = res$t
  names(tstat) = res$symbol
  tstat = tstat[!is.na(names(tstat))]

  index = ids2indices(geneSetsCombined, names(tstat))

  cameraPR( tstat, index )
}

plot_compare_enrichment = function( res1, res2, gset, col=c("grey60", "royalblue1"),maxValue=30){

  res1 = res1[rownames(res1) %in% gset,]
  res2 = res2[rownames(res2) %in% gset,]

  res1$dataset = 'dupCor'
  res2$dataset = 'dream'

  res1$name = rownames(res1)
  res2$name = rownames(res2)

  res = data.table(rbind(res1, res2))
  # res$name = substring(rownames(res), 1, 30)
  res$dataset = factor(res$dataset, c("dupCor", "dream"))

  res$name = substring( res$name, 1,60)

  df = res[,min(FDR),by='name']

  res$name = factor(res$name, df[,name[order(V1, decreasing=TRUE)]])

  res = res[order(res$name),]

  ggplot(res, aes(name, -log10(FDR), fill=dataset)) + geom_bar(stat='identity', position=position_dodge()) + coord_flip() + theme_bw(12) + theme(aspect.ratio=2, plot.title = element_text(hjust = 0.5), legend.position="none") + ylab(bquote(-log[10]~FDR)) + xlab("Gene sets") + scale_fill_manual(values=col) + geom_hline(yintercept=-log10(0.05), linetype=2) + geom_hline(yintercept=-log10(0.01), linetype=3) + scale_y_continuous(expand = c(0, 0), lim=c(0, maxValue))
}

formTable = function(x,digits, stop=40){
  if(missing(digits)){
    digits = rep(2, ncol(x))
  }
  if( length(digits) != ncol(x)){
    stop("Lengths don't match")
  }
  x = data.frame(x)
  ret = sapply(1:ncol(x), function(i){
    format(x[,i,drop=FALSE],digits=digits[i])})
  ret = do.call("cbind", ret)
  rownames(ret) = rownames(x)
  ret[,1] = substring(ret[,1], 1, stop)
  rownames(ret) = substring(rownames(ret), 1, stop)
  ret
}
```


```{r vp}
form = ~ (1|Region:bbscore) + (1|individualIdentifier) + (1|Run) + (1|LIMS) + (1|Region) + (1|`Sex (final)`) + RIN + rRNA.rate + PMI + Age + mappingRate + Apo1 + Apo2 
vp = fitExtractVarPartModel( vobj, form, metadata)
```

```{r vp.plot}
plotVarPart( sortCols( vp ) )      
```


# Compare expression with Braak stage
```{r limma_braak, echo=FALSE, message=FALSE}
n_samples = table(metadata$individualIdentifier)

# idx = with(metadata, !is.na(DiseaseClass) & (!is.na(Apo1)) &(!is.na(bbscore)) &(individualIdentifier %in%  names(n_samples[n_samples==4])) )
idx = rep(TRUE, nrow(metadata))

design = model.matrix( ~ 0 + Region + Region:bbscore + Run + LIMS + `Sex (final)` + RIN + rRNA.rate + PMI + Age + Apo1  + Apo2, metadata[idx,])
 
# Estimate linear mixed model with a single variance component
# Fit the model for each gene, 
dupcor <- duplicateCorrelation(vobj[,idx],design,block=metadata$individualIdentifier[idx])

# But this step uses only the genome-wide average for the random effect
fitDupCor <- lmFit(vobj[,idx], design, block=metadata$individualIdentifier[idx], correlation=dupcor$consensus)

Lc = matrix(0, ncol=ncol(design)) 
colnames(Lc) = colnames(design)
Lc[length(Lc):(length(Lc)-3)] = 1

fitDupCor = contrasts.fit( fitDupCor, t(Lc))
 
# Fit Empirical Bayes for moderated t-statistics
fitDupCor <- eBayes( fitDupCor )
```

```{r dream_analysis_braak, echo=FALSE, message=FALSE}
form = ~ 0 + Region + Region:bbscore + (1|individualIdentifier) + (1|Run) + (1|LIMS)  + `Sex (final)` + scale(RIN) + scale(rRNA.rate) + scale(PMI) + scale(Age) + scale(mappingRate) + Apo1  + Apo2 
 
# Get the contrast matrix for the hypothesis test
L = getContrast( vobj[,idx], form, metadata[idx,], "RegionBM44:bbscore")
L['RegionBM10:bbscore'] = 1
L['RegionBM22:bbscore'] = 1 
L['RegionBM36:bbscore'] = 1 

# Fit the dream model on each gene
# Apply the contrast matrix L for the hypothesis test  
# By default, uses the Satterthwaite approximation for the hypothesis test
fitmm = dream( vobj[,idx], form, metadata[idx,], L)

fitmm = eBayes( fitmm )
```

# Compare -log10 p from dream and duplicateCorrelation
```{r compare_braak}
p1 = topTable(fitDupCor, number=Inf, sort.by="none")$P.Value
p2 = topTable(fitmm, number=Inf, sort.by="none")$P.Value
      
plotCompareP( p1, p2, vp$individualIdentifier, dupcor$consensus)
```


# Top 20 categories from each analysis
```{r camera_enrich_braak}
res = topTable(fitDupCor, number=1e7) 
enrich_dupCor_camera = get_enrich_order( res ) 
   
res = topTable(fitmm, number=1e7)  
enrich_dream_camera = get_enrich_order( res )   
```
 
```{r compare_enrichment_braak, fig.width=12}
res1 = enrich_dupCor_camera 
    
res2 = enrich_dream_camera     

col = c('#00b6eb', '#a58affff') 
 
maxValue = -log10(min(c(min(res1$FDR), min(res2$FDR)))) * 1.03

# plot_compare_enrichment( res1, res2, rownames(res1)[1:20], col, maxValue)
# plot_compare_enrichment( res1, res2, rownames(res2)[1:20], col, maxValue)  

geneSetNames = unique(c(rownames(res1)[1:20], rownames(res2)[1:20]))
plot_compare_enrichment( res1, res2, geneSetNames, col, maxValue)  
```

## Plot for publication
```{r plot_enrichment_publish}
gset = c(
'c2.BURTON_ADIPOGENESIS_PEAK_AT_2HR',
'c2.UZONYI_RESPONSE_TO_LEUKOTRIENE_AND_THROMBIN',
'h.HALLMARK_HYPOXIA',
'c2.PHONG_TNF_TARGETS_UP',
'c2.BURTON_ADIPOGENESIS_1',
'c7.GSE9988_ANTI_TREM1_AND_LPS_VS_CTRL_TREATED_MONOCYTES_DN',
'h.HALLMARK_TNFA_SIGNALING_VIA_NFKB',
'GSE9988_LPS_VS_VEHICLE_TREATED_MONOCYTE_DN')

fig = plot_compare_enrichment( res1, res2, gset, col, maxValue) 
fig + theme(aspect.ratio=length(gset)/3) 
```

# Number of differentially expressed genes from DupCor 
```{r}
sum(topTable(fitDupCor, number=Inf)$adj.P.Val < 0.05)
```

# Number of differentially expressed genes from dream 
```{r}
sum(topTable(fitmm, number=Inf)$adj.P.Val < 0.05)
```


# Variance fractions of expression plots for two example genes
### TUBB2B: p-values are very different between the two methods
```{r plotStratify.TUB}
vp2 = with( vp, data.frame(Batch = LIMS + Run, Brain_region = Region, Individual = individualIdentifier + Apo1 + Apo2 + `Sex (final)`, QC_metrics =  RIN + rRNA.rate + PMI + mappingRate, Age = Age, Braak_stage = `Region:bbscore`,  Residuals = Residuals)) 
vp2 = sortCols( vp2 )
rownames(vp2) = rownames(vp)

ensGene = rownames(topTable(fitDupCor))[1]

GE = data.frame( Expression = vobj$E[ensGene,], Indiv = metadata$individualIdentifier,  bbscore = metadata$bbscore)
     
fig1 = plotStratify( Expression ~ Indiv, GE, colorBy = GE$bbscore, main= ensGene, legend=TRUE) + theme(aspect.ratio=1, legend.position="right") + scale_fill_continuous(name = "Braak stage") + scale_color_continuous(name = "Braak stage")

fig2 = plotPercentBars( vp2[ensGene,]) + theme(aspect.ratio=.1) 

grid.arrange( fig1, fig2, ncol=1, layout_matrix = t(t(c(1, .1))) )
```









